Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-09-23 bmc_template/figures.tex
ready to submit

\documentclass{article}\usepackage[]{graphicx}\usepackage{xcolor}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/r_msdlist-1}

@@

\section{sup}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot2-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot2-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot14-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot14-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot15-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot15-1}

>>>>

\documentclass{article}\usepackage[]{graphicx}\usepackage{xcolor}
\usepackage{lineno}
\linenumbers

%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\renewcommand{\thetable}{A\arabic{table}}

\renewcommand{\thefigure}{A\arabic{figure}}
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\section*{Additional files}

  In the main text of this manuscript, we showed the fitting and simulating results for Cat 1. Here, we have the adjusted BIC emission distribution comparisons, adjusted BIC across all FMM and HMM models, diurnal step lengths and autocorrelation plots for Cat 2, 14, and 15. Table A1 gives the name of panther ID (Cat ID) in Institutional Repository at the University of Florida (IR@UF), \cite{kerk2015hidden}, and number of observations for each panther.

\begin{table}[h!]
\caption{Cat ID and number of observations; ID numbers are given matching
those shown by van de Kerk et al. 2015 and those 
in the data located at the UF Institutional repository (IR@UF).}
\begin{tabular}{cccc}
  \hline
  & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
  & 130 & 1 & 10286\\
  & 131 & 2  & 9458\\
  & 48 & 14  & 14645\\
  & 94  & 15   & 10250\\ \hline
\end{tabular}
\end{table}

\newpage

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/BICred_plot2-1} \caption{Cat 2: Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and heterogeneous transition HMMs (right panel) with different emission complexities on panther data. Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution). Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution.}
\label{fig:BICred_plot2}
\end{figure}

\clearpage
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons2-1} \caption{Cat 2: Comparison of adjusted BIC for all HMM transition complexities. The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM. The right panel shows HMMs with different temporal transitions. Dashed lines represents FMMs and solid lines represents HMMs.}\label{fig:adj_BIC_comparisons2}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time2-1} \caption{Cat 2: Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}\label{fig:avg_step_length_by_time2}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/acf_plot2-1} \caption{Cat 2: Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
\end{figure}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/BICred_plot14-1} \caption{This figure mataches Figure A1, but for Cat 14.}\label{fig:BICred_plot14}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons14-1} \caption{This figure mataches Figure A2, but for Cat 14.}\label{fig:adj_BIC_comparisons14}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time14-1} \caption{This figure mataches Figure A3, but for Cat 14.}\label{fig:avg_step_length_by_time14}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/acf_plot14-1} \caption{This figure mataches Figure A4, but for Cat 14.}\label{fig:acf_plot14}
\end{figure}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/BICred_plot15-1} \caption{This figure mataches Figure A1, but for Cat 15.}\label{fig:BICred_plot15}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons15-1} \caption{This figure mataches Figure A2, but for Cat 15.}\label{fig:adj_BIC_comparisons15}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time15-1} \caption{This figure mataches Figure A3, but for Cat 15.}\label{fig:avg_step_length_by_time15}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/acf_plot15-1} \caption{This figure mataches Figure A4, but for Cat 15.}\label{fig:acf_plot15}
\end{figure}

@@

\end{knitrout}
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{panthers}

### 2016-09-23 bmc_template/manuscript.tex
ready to submit

%% TO DO
% author references (maybe fix programmatically?)
% acknowledgments

@@

van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  To check that this simplification
does not distort our conclusions
we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 2).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

@@

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/BICred_plot-1}
  \caption{Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and and heterogeneous transition HMMs (right panel) with different emission complexities on panther data. Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution). Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution. }
      \end{figure}

The BIC-optimal number of states for time-homogeneous models is
consistent with van de Kerk et al.'s \cite{kerk2015hidden}
results. For time-homogeneous models, the Weibull-wrapped-Cauchy
\cite{kerk2015hidden}, Weibull-von Mises, and log Normal without
turning angles all identify the same BIC-optimal number of
states. While the number of states identified
by homogeneous-HMM models varies according to the emissions
distributions chosen, ranging from $n=5$ for Weibull
steps alone to $n=7$ for the log Normal-von Mises emissions
model, the number of states identified
by heterogeneous-HMM models
is consistent among emissions models ($n=5$: Figure 2).

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
  \caption{Comparison of adjusted BIC for all HMM transition complexities. The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM. The right panel shows HMMs with different temporal transitions. Dashed lines represents FMMs and solid lines represents HMMs.}
      \end{figure}

Models with temporal heterogeneity provide better fits to the data
(lower BIC) than homogeneous models in both FMM and HMM frameworks,
but time-homogeneous HMMs are better than FMMs with sinusoidal
temporal heterogeneity (Figure 3). Turning to the temporally
heterogeneous HMMs (Figure 3, right panel), we see that the model with
different transition probabilities for each hour of the day (HMM +
THhourly) is overparameterized; it underperforms homogeneous HMM with
even 3 states, and gets much worse with 4 states. The multiple-block
model gives approximately the same BIC as the homogeneous HMM,
although it gives the BIC-optimal number of states as 4, in contrast
to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal
models are the best models tested by far; they both give the
BIC-optimal number of states as 5, and they have similar goodness of
fit.  However, the similarity between the quadratic 
and sinusoidal models may be overstated in Figure~3 due to the very large
variation in BIC (over thousands of units) across the full range of
models; the best-fit sinusoidal ($n=5$) model is approximately 80 BIC units
better than the best quadratic model (also $n=5$), which
would normally be interpreted as an enormous improvement in goodness of
fit (both models have 90 parameters).

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/avg_step_length_by_time-1}
  \caption{Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
      \end{figure}

The average hourly step lengths from the observed panther data exhibit
a clear diurnal pattern (Figure 4). As expected, temporally
homogeneous models (whether FMM or HMM) predict the same mean step
length regardless of time of day, failing to capture the diurnal
activity cycle. All of the models incorporating temporal
heterogeneity, including the temporally heterogeneous FMM, can capture
the observed patterns. However, the block model does markedly worse
than the other temporal models (changing the block definitions might
help by re-clustering/grouping different hours or increasing the
number of blocks), and the (overparameterized) hourly model does
better than any other model at capturing the early-evening peak (but
worse at capturing the mid-day trough). We also included average
hourly step lengths from three-state temporally homogeneous HMM
Viterbi prediction to illustrate within sample predictions can capture
the diurnal patterns, but fail to capture out of sample predictions.

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/acf_plot-1}
  \caption{Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
      \end{figure}

Like the diurnal pattern (Figure 4),
the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMMs, by definition, produce no autocorrelation (neither short- nor long-term autocorrelation). FMMs without temporal heterogeneity, although they capture the diurnal pattern well, underpredict the degree of short-term autocorrelation.

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{State identification of BIC-optimal HMMs for all panthers. Mean (x-axis) and standard deviation (y-axis) of step length by state $\textrm{log}_{10} m$ for BIC-optimal homogeneous HMMs (red line) and heterogeneous HMMs with sinusoidal transition (blue line). Short movement states (step-length \textless $10^{1.5}$m) are circled together.   
  }
      \end{figure}

The estimated emission parameter values
(mean and standard deviation of the step length in each state)
are similar, for both homogeneous and heterogeneous
models, across all cats (Figure 6 shows a subset of cats). 
In general, the states with longer mean step lengths are
similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at \url{http://ufdc.ufl.edu//IR00004241/00001}.

@@

@@

\section*{Tables}
\begin{table}[h!]
\caption{Cat ID and number of observations; ID numbers are given matching
those shown by van de Kerk et al. 2015 and those 
in the data located at the UF Institutional repository (IR@UF).}
\begin{tabular}{cccc}
  \hline
  & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
  & 130 & 1 & 10286\\
  & 131 & 2  & 9458\\
  & 48 & 14  & 14645\\
  & 94  & 15   & 10250\\ \hline
\end{tabular}
\end{table}

\end{backmatter}
\end{document}

>>>>

@@

van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  To check that this simplification
does not distort our conclusions
we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 1).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

@@

The BIC-optimal number of states for time-homogeneous models is
consistent with van de Kerk et al.'s \cite{kerk2015hidden}
results. For time-homogeneous models, the Weibull-wrapped-Cauchy
\cite{kerk2015hidden}, Weibull-von Mises, and log Normal without
turning angles all identify the same BIC-optimal number of
states. While the number of states identified
by homogeneous-HMM models varies according to the emissions
distributions chosen, ranging from $n=5$ for Weibull
steps alone to $n=7$ for the log Normal-von Mises emissions
model, the number of states identified
by heterogeneous-HMM models
is consistent among emissions models ($n=5$: Figure 1).

Models with temporal heterogeneity provide better fits to the data
(lower BIC) than homogeneous models in both FMM and HMM frameworks,
but time-homogeneous HMMs are better than FMMs with sinusoidal
temporal heterogeneity (Figure 2). Turning to the temporally
heterogeneous HMMs (Figure 2, right panel), we see that the model with
different transition probabilities for each hour of the day (HMM +
THhourly) is overparameterized; it underperforms homogeneous HMM with
even 3 states, and gets much worse with 4 states. The multiple-block
model gives approximately the same BIC as the homogeneous HMM,
although it gives the BIC-optimal number of states as 4, in contrast
to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal
models are the best models tested by far; they both give the
BIC-optimal number of states as 5, and they have similar goodness of
fit.  However, the similarity between the quadratic 
and sinusoidal models may be overstated in Figure~2 due to the very large
variation in BIC (over thousands of units) across the full range of
models; the best-fit sinusoidal ($n=5$) model is approximately 80 BIC units
better than the best quadratic model (also $n=5$), which
would normally be interpreted as an enormous improvement in goodness of
fit (both models have 90 parameters).

The average hourly step lengths from the observed panther data exhibit
a clear diurnal pattern (Figure 3). As expected, temporally
homogeneous models (whether FMM or HMM) predict the same mean step
length regardless of time of day, failing to capture the diurnal
activity cycle. All of the models incorporating temporal
heterogeneity, including the temporally heterogeneous FMM, can capture
the observed patterns. However, the block model does markedly worse
than the other temporal models (changing the block definitions might
help by re-clustering/grouping different hours or increasing the
number of blocks), and the (overparameterized) hourly model does
better than any other model at capturing the early-evening peak (but
worse at capturing the mid-day trough). We also included average
hourly step lengths from three-state temporally homogeneous HMM
Viterbi prediction to illustrate within sample predictions can capture
the diurnal patterns, but fail to capture out of sample predictions.

Like the diurnal pattern (Figure 3),
the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 4) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMMs, by definition, produce no autocorrelation (neither short- nor long-term autocorrelation). FMMs without temporal heterogeneity, although they capture the diurnal pattern well, underpredict the degree of short-term autocorrelation.

The estimated emission parameter values
(mean and standard deviation of the step length in each state)
are similar, for both homogeneous and heterogeneous
models, across all cats (Figure 5 shows a subset of cats). 
In general, the states with longer mean step lengths are
similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

\section*{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at \url{http://ufdc.ufl.edu//IR00004241/00001}.

@@

\end{backmatter}

@@

\section*{Tables}

\clearpage

\section*{Figures}
  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/Figure-1}
  \caption{Comparison of BIC-optimal state predictions for homogeneous transition HMMs (left panel) and heterogeneous transition HMMs (right panel) with different emission complexities on panther data. Solid line represents univariate response/emission HMMs (without turning angles) and dotted line represents multivariate response HMMs (including turning angles with von-Mises distribution). Red lines represents log-normal step-length distribution and blue lines represents Weibull step-length distribution. }
      \end{figure}

\clearpage

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/Figure-2}
  \caption{Comparison of adjusted BIC for all HMM transition complexities. The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM. The right panel shows HMMs with different temporal transitions. Dashed lines represents FMMs and solid lines represents HMMs.}
      \end{figure}

\clearpage

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/Figure-3}
  \caption{Comparison of average step-length by time of day out of sample predictions for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
      \end{figure}

\clearpage

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/Figure-4}
  \caption{Comparison of out of sample predictions autocorrelations for BIC-optimal state FMMs (dashed line) and HMMs (solid lines) of different transition complexities. Gray highlight indicate the observed average step length from panther and "V" points represents Viterbi predictions (within sample) of a three-state homogeneous transition HMM.}
      \end{figure}

\clearpage

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/Figure-5}
  \caption{State identification of BIC-optimal HMMs for all panthers. Mean (x-axis) and standard deviation (y-axis) of step length by state $\textrm{log}_{10} m$ for BIC-optimal homogeneous HMMs (red line) and heterogeneous HMMs with sinusoidal transition (blue line). Short movement states (step-length \textless $10^{1.5}$m) are circled together.   
  }
      \end{figure}

\end{document}
